---
title: "Practicum1"
author: "Thomas Faria, Caitlin Kirkpatrick, Sairah Shir"
date: "2024-06-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages for Practicum1
library(tidyverse)
library(devtools)
```

## Part 1: Questions

### Question 1

```{r, q1}
# Define dataframe
doc_df <- data.frame(
  doctor_type = c("PCP", "Psychiatrist", "Surgeon", "Anesthesia"),
  doctor_lastname = c("Smith", "Dame", "Jones", "Zayas"),
  location = c("MA", "ME", "NH", "VT"),
  AVG_Rating = c("7", "9", "8", "9")
)
print(doc_df)
```

### Question 2

```{r, q2}
# Index with brackets
doc_df[1, 2]
doc_df[2:4, ]
doc_df[, 4]
```

### Question 3

```{r, q3}
library(ggplot2)
head(mtcars)


ggplot(mtcars,aes(x=wt, y=mpg, color=factor(cyl))) + geom_point() + labs(x="Weight", y="MPG", title = "Relationship between MPG and Weight", caption = "This graph looks at the relationship between MPG and weight using cylinders as the color scheme.")

```

### Question 4

```{r}
# Examine variable summary stats
summary(mtcars)

mpg_weight_cor <- mtcars %>%
select(wt, mpg) %>%
drop_na() %>%
summarize(correlation = cor(wt, mpg))
mpg_weight_cor
```

* It is known that the weight of a vehicle plays a key role in how fuel efficient it is, so both wt and mpg were selected to identify a potential correlation
* The Pearson coefficient is a descriptive statistic that reveals the linear correlation between two variables
* A value between 0 and 1 shows how strong the correlation is, with 1 indicating a strong correlation and 0 an absence of correlation
* The sign of the value indicates the type of correlation (positive or negative)
* In this case, the coefficient value returned is approximately -0.87, which indicates a strong, negative correlation
  * As vehicle weight increases, fuel efficiency (mpg) decreases

---

## Part 2: Practicum Tasks

#### Load data from provided URL

```{r, load_data}

# Documentation from data.world recommends package installation directly from Github
devtools::install_github("datadotworld/data.world-r", build_vignettes = TRUE)

# Load the requisite API token obtained from data.world advanced settings (Thomas's account)
# Original code: token <- readLines('~/RStudioProjects/Summer24_DA5020_Group7_Practicum1/API_token')
# Changed
token <- readLines('API_token')
saved_cfg <- data.world::save_config(token)
data.world::set_config(saved_cfg)

# From data.world R and RStudio integration:
library("data.world")
sql_stmt <- data.world::qry_sql("SELECT * FROM chemical_dependence_treatment_program_admissions_beginning_2007_1")
query_results_df <- data.world::query(
  sql_stmt, "https://data.world/data-ny-gov/ngbt-9rwf")

```

#### Initial evaluation of dataset

-   Determine necessary preparation steps and perform them

-   Discuss distribution, outliers, and prepare summary stats

```{r, q4}

# Evaluate data distribution, outliers, and prepare summary stats

# Reassign query results to more descriptive variable
admissions_data <- query_results_df

# Overview data
glimpse(admissions_data)
summary(admissions_data)

# Visualize outliers in the admissions column
ggplot(admissions_data) +
  labs(title = "Admissions Data Distribution") +
  geom_histogram(mapping = aes(x = admissions), bindwidth = 5) + 
  labs(title = "Distribution of Admissions Values",
       x = "Admissions",
       y = "Count",
       caption = "A histogram showing the distribution of numerical admissions data")

# Zoom in on low occuring values
ggplot(admissions_data) +
  geom_histogram(mapping = aes(x = admissions), binwidth = 5) +
  labs(title = "Distribution of Admissions Values (< 8 occurences)",
       x = "Admissions",
       y = "Count",
       caption = "Instances of less-occuring admissions values") +
  coord_cartesian(ylim = c(0,8))

# Designate all columns except for year, admissions, and county as categorical (using as.factor())
admissions_data_factors <- admissions_data %>% 
  mutate(
    program_category = as.factor(program_category),
    service_type = as.factor(service_type),
    age_group = as.factor(age_group),
    primary_substance_group = as.factor(primary_substance_group)
  )

# Compute summaries per factor
program_category_summary <- admissions_data_factors %>% 
  group_by(program_category) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  ) 
print(program_category_summary)

service_type_summary <- admissions_data_factors %>% 
  group_by(service_type) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
print(service_type_summary)

age_group_summary <- admissions_data_factors %>% 
  group_by(age_group) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
print(age_group_summary)

primary_substance_group_summary <- admissions_data_factors %>% 
  group_by(primary_substance_group) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
print(primary_substance_group_summary)
```

```{r}
# Compute outliers for admissions
admissions_outliers <- admissions_data_factors %>%
  mutate(
    mean_admissions = mean(admissions, na.rm = TRUE),
    sd_admissions = sd(admissions, na.rm = TRUE)
  ) %>%
  # Relative to the mean, any values on the lower or upper bounds that are 3 times the standard deviation are filtered
  filter(admissions < mean_admissions - 3 * sd_admissions | admissions > mean_admissions + 3 * sd_admissions) %>%
  select(admissions)
admissions_outliers

# Remove outliers from dataset
rmv_admissions_outliers <- admissions_data_factors %>%
  mutate(
    mean_admissions = mean(admissions, na.rm = TRUE),
    sd_admissions = sd(admissions, na.rm = TRUE)
  ) %>%
  filter(!(admissions < mean_admissions - 3 * sd_admissions | admissions > mean_admissions + 3 * sd_admissions))
# Note subtracted outliers from new dataframe
str(admissions_data_factors)
str(rmv_admissions_outliers$admissions)
```

* It may not be significant that admissions numbers are higher than usual at any given point in time
* Outliers are presented in one dataframe as being high values 

#### Restructure data into appropriate tibbles

```{r county}
# Read in .csv created from https://www.dot.ny.gov/main/business-center/engineering/specifications/local-specifications/county-code as reference
county_codes <- read_csv("county_codes.csv") 
county_codes
```

```{r county1}
county <- admissions_data_factors %>%
  select(county_of_program_location) %>%
  distinct() %>%
  # Join codes with respective counties from county_codes
  left_join(county_codes, by = "county_of_program_location") %>%
  mutate(county_code = case_when(
      # Tagging counties with first two characters and "-NYC"
      county_of_program_location %in% c("Bronx", "Queens", "Kings") ~ paste(str_to_upper(str_sub(county_of_program_location, 1, 2)), "-NYC", sep = ""),
      county_of_program_location == "New York" ~ "NYC",
      # Handle to not treat as NA value, changed code to NS instead
      county_of_program_location == "Nassau" ~ "NS",
      # Handle to not treat as NA value, manually assigned SL 
      county_of_program_location == "St Lawrence" ~ "SL",
      TRUE ~ county_code 
    )
  )
# Note: The county "Hamilton" is included in the county_codes csv, but it is not found in the admissions_data dataframe
county
```

```{r, program_category}

# Define abbreviations for recoding
program_category_index <- c(
  "Crisis" = "C",
  "Inpatient" = "I",
  "Opioid Treatment Program" = "OTP",
  "Outpatient" = "O",
  "Residential" = "R"
)

# Ceate new column called program_code
admissions_data_coded <- admissions_data_factors %>%
  mutate(program_code = recode(program_category, 
       "Crisis" = "C",
       "Inpatient" = "I",
       "Opioid Treatment Program" = "OTP",
       "Outpatient" = "O",
       "Residential" = "R"))

# Create program_category tibble and recode based on the index directly
program_category_df <- admissions_data_coded %>%
  distinct(program_category, .keep_all = TRUE) %>%
  select(program_code, program_category)

program_category_df

```

```{r, primary_substance_group}
# Sairah
# Define the index for recoding
primary_substance_group_index <- c(
  "Alcohol" = "A",
  "All Others" = "AO",
  "Cocaine incl Crack" = "CC",
  "Heroin" = "H",
  "Marijuana incl Hashish" = "MH",
  "Other Opioids" = "OO"
)

# Update admissions_data_coded with new column called substance_code
admissions_data_coded <- admissions_data_coded %>%
  mutate(substance_code = recode(primary_substance_group, 
        "Alcohol" = "A", 
        "All Others" = "AO",
        "Cocaine incl Crack" = "CC",
        "Heroin" = "H",
        "Marijuana incl Hashish" = "MH",
        "Other Opioids" = "OO"))

# Create Primary Substance Group tibble and recode based on the index directly
primary_substance_group_df <- admissions_data_coded %>%
  distinct(primary_substance_group, .keep_all = TRUE) %>%
  select(substance_code, primary_substance_group)
 
primary_substance_group_df

```

```{r, admissions_data}
# Thomas

# Join county_code data onto main tibble using a full_join by county name
admissions_data_coded_joined <- admissions_data_coded %>% 
  full_join(county, by = "county_of_program_location")

# Final tibble: admissions_data_df
admissions_data_df <- admissions_data_coded_joined %>% 
  select(
    year,
    county_code,
    program_code,
    service_type,
    age_group,
    substance_code,
    admissions
  )

```

#### Define annualAdmissions()

-   Function should derive the total \# of reported admissions per year for the entire state of NY and display these results on a line graph

-   Annotate to show year with highest admissions

-   Explain results

```{r, define_function}
# Thomas -- NEEDS WORK

# This function uses aggregate() to sum the total admissions for every year in the admissions_data_df
# The max point is computed from the aggregated tibble and held for later reference on the graph
# A line graph is prepared using ggplot2 with appropriate labeling

annualAdmissions <- function() {
  # Get aggregated data as its own tibble for easy ref
  total_admissions <- aggregate(admissions_data_df$admissions, 
            by = list(year = admissions_data_df$year), 
            sum) %>% 
    rename(total = x)
  
  # Get maximum
  max_point <- total_admissions[which.max(total_admissions$total), ]
  
  # Plot a line graph
  total_admissions %>% 
    ggplot(mapping = aes(x = year, y = total)) + 
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 2007:2017) +
    scale_y_continuous(limits = c(270000, 320000)) +
    labs(title = "Total admissions per year, all of New York State",
         x = "Total admissions",
         y = "Year") + 
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    ) + 
    annotate("text",
             x = max_point$year,
             y = max_point$total,
             label = paste("Max. admissions:", max_point$total, "in", max_point$year),
             vjust = -1)
}

annualAdmissions()

```

#### Analyze % of admissions by county

-   Visualize top 5 counties using a bar chart

-   Explain results

```{r, admissions_by_county}
# Sairah

#total number of admission in the NYS
total= sum(admissions_data_df$admissions)

#calculate percentage of admissions in each county 
percentage_admissions <- admissions_data_df %>% 
  select(county_code, admissions) %>%
  group_by(county_code) %>% 
  summarize(percentage=((sum(admissions)/total) * 100))

top_5_admission_counties <- percentage_admissions %>% 
  slice_max(n=5, percentage)

print(top_5_admission_counties)

library(ggplot2)

ggplot(top_5_admission_counties, aes(x=county_code, y = percentage)) + geom_col(fill = "blue") + 
    labs(x = "County", y= "Percentage (%)", title = "Top 5 Admission Percentage by Counties", caption = "This bar chart shows the top five counties in New York City that \nhad the highest percentage of admissions to the chemical dependence treatment program.") + 
  scale_x_discrete(labels = c("NYC" = "New York City", "QU-NYC" = "Queens", "BR-NYC" = "Bronx", "KI-NYC" = "Kings", "SU" = "Suffolk"))

```
*To find the percentage of admissions in counties, first the total sum of admissions was needed. Then the admissions data dataframe was used and the county code and admission columns were selected. Then grouped by the county code and summarized by calculating the percentage. The percentage admission was used to find the top fix counties using slice_max function and then put into a bar chart 
* In the bar chart it shows that the counties with the most admissions were the most populated counties in New York. 

#### Extract various "Rehab" facilities information

-   Use a regex to match all facilities that include the word rehab, rehabilitation, etc.

-   Using filtered data, identify the most prominent (common) substance related to admission for each age group

-   Visualize and explain results

```{r, rehab_regex}
rehab_df <- admissions_data_factors %>%
  # Only show rehabilitation services
  filter(str_detect(service_type, regex("Rehab|Rehabilitation", ignore_case = TRUE))) %>%
  select(service_type, age_group, primary_substance_group, admissions)
rehab_df

top_substance_df <- rehab_df %>%
  # Only interested in these combinations
  group_by(service_type, age_group, primary_substance_group) %>%
  # Take count to show how many admissions exist for each substance in each age group
  summarize(substance_count = sum(admissions)) %>%
  # Limit to age group 
  group_by(age_group) %>%
  # Filter for the substances with the highest count
  filter(substance_count == max(substance_count)) %>%
  # Show relevant columns
  select(service_type, age_group, primary_substance_group, substance_count)
top_substance_df
```

* To identify the most prominent substance used in each age group, we first define a regular expression in a new dataframe that filters all services containing "Rehab" or "Rehabilitation" in the name
* A separate dataframe is then defined to find the top substance per age group
  * The data is grouped by age_group and primary_substance_group because we are only
  interested in analyses in the context of these variables paired together

```{r}
ggplot(top_substance_df, aes(primary_substance_group, substance_count, fill = age_group)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(
    name = "Substance",
    labels = c(
      "All Others" = "Other",
      "Cocaine incl Crack" = "Cocaine (Includes Crack)",
      "Marijuana incl Hashish" = "Marijuana (Includes Hashish)"
    )
  ) +
  scale_fill_discrete(
    name = "Age Group",
    labels = c(
      "18 thru 24" = "18-24",
      "25 thru 34" = "25-34",
      "35 thru 44" = "35-44",
      "45 thru 54" = "45-54",
      "55 and Older" = "55+",
      "Under 18" = "Under 18"
    )
  ) +
  labs(
    y = "Number of Admissions",
    title = "Top substances used per age group",
    caption = "Substances grouped by top use per age group with the number of\nadmissions on they-axis and the type of substance on the x-axis."
  ) +
  theme(
    axis.title = element_text(face = "bold", color = "black")
  ) +
  theme_minimal()
```


```{r}
all_substance_df <- rehab_df %>%
  group_by(service_type, age_group, primary_substance_group) %>%
  # Take count to show how many admissions exist for each substance in each age group
  summarize(substance_count = sum(admissions)) %>%
  # Limit to age group 
  group_by(age_group) %>%
  # Filter for the substances with the highest count
  mutate(substance_count == max(substance_count)) %>%
  # Show relevant columns
  select(service_type, age_group, primary_substance_group, substance_count)
all_substance_df
```

```{r all substance}
ggplot(all_substance_df, aes(age_group, substance_count, fill = primary_substance_group)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Age Group",
    y = "Number of Admissions",
    title = "All substances used per age group",
    caption = "Stacked breakdown of all substances used per age\ngroup with the number of admissions on the\ny-axis and the age groups on the x-axis."
  ) +
  scale_fill_discrete(
    name = "Substance",
    labels = c(
      "All Others" = "Other",
      "Cocaine incl Crack" = "Cocaine (Includes Crack)",
      "Marijuana incl Hashish" = "Marijuana (Includes Hashish)"
    )
  ) +
  scale_x_discrete(
    labels = c(
      "18 thru 24" = "18-24",
      "25 thru 34" = "25-34",
      "35 thru 44" = "35-44",
      "45 thru 54" = "45-54",
      "55 and Older" = "55+",
      "Under 18" = "Under 18"
    )
  ) +
  theme(
    axis.title = element_text(face = "bold", color = "black")
  ) +
  theme_minimal()
```

```{r service}
ggplot(all_substance_df, aes(age_group, substance_count, fill = service_type)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Age Group",
    y = "Number of Admissions",
    title = "Service used per age group",
    caption = "Stacked breakdown of services used for\nadmissions per age group with number of\nadmissions on the y-axis and age group on the y-axis."
  ) +
  scale_fill_discrete(
    name = "Service Type",
    labels = c(
      "Inpatient Rehabilitation" = "Inpatient",
      "Outpatient Rehabilitation" = "Outpatient",
      "Res Rehab for Youth" = "Youth Residential",
      "Residential Rehabilitation" = "Residential",
      "Specialized Services OP Rehab" = "Specialized Services OP"
    )
  ) +
  scale_x_discrete(
    labels = c(
      "18 thru 24" = "18-24",
      "25 thru 34" = "25-34",
      "35 thru 44" = "35-44",
      "45 thru 54" = "45-54",
      "55 and Older" = "55+",
      "Under 18" = "Under 18"
    )
  ) +
  theme(
    axis.title = element_text(face = "bold", color = "black")
  ) +
  theme_minimal()
```
