---
title: "Practicum1"
author: "Thomas Faria"
date: "2024-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages for Practicum1
library(tidyverse)
library(devtools)

```

## Part 1: Questions

### Question 1

```{r, q1}

```

### Question 2

```{r, q2}

```

### Question 3

```{r, q3}

```

### Question 4

```{r, q4}

```

## Part 2: Practicum Tasks

#### Load data from provided URL

```{r, load_data}

# Documentation from data.world recommends package installation directly from Github
devtools::install_github("datadotworld/data.world-r", build_vignettes = TRUE)

# Load the requisite API token obtained from data.world advanced settings (Thomas's account)
token <- readLines('~/RStudioProjects/Summer24_DA5020_Group7_Practicum1/API_token')
saved_cfg <- data.world::save_config(token)
data.world::set_config(saved_cfg)

# From data.world R and RStudio integration:
library("data.world")
sql_stmt <- data.world::qry_sql("SELECT * FROM chemical_dependence_treatment_program_admissions_beginning_2007_1")
query_results_df <- data.world::query(
  sql_stmt, "https://data.world/data-ny-gov/ngbt-9rwf")

```

#### Initial evaluation of dataset

-   Determine necessary preparation steps and perform them

-   Discuss distribution, outliers, and prepare summary stats

```{r, prep_data}

```

#### Restructure data into appropriate tibbles

```{r, county}
# Caitlin
```

```{r, program_category}
# Sairah
```

```{r, primary_substance_group}
# Sairah
```

```{r, admissions_data}
# Thomas
```

#### Define annualAdmissions()

-   Function should derive the total \# of reported admissions per year for the entire state of NY and display these results on a line graph

-   Annotate to show year with highest admissions

-   Explain results

```{r, define_function}

```

#### Analyze % of admissions by county

-   Visualize top 5 counties using a bar chart

-   Explain results

```{r, admissions_by_county}

```

#### Extract various "Rehab" facilities information

-   Use a regex to match all facilities that include the word rehab, rehabilitation, etc.

-   Using filtered data, identify the most prominent (common) substance related to admission for each age group

-   Visualize and explain results

```{r, rehab_regex}

```
