---
title: "Practicum1"
author: "Thomas Faria"
date: "2024-06-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required packages for Practicum1
library(tidyverse)
library(devtools)
```

## Part 1: Questions

### Question 1

```{r, q1}
doc_df <- data.frame(
  doctor_type = c("PCP", "Psychiatrist", "Surgeon", "Anesthesia"),
  doctor_lastname = c("Smith", "Dame", "Jones", "Zayas"),
  location = c("MA", "ME", "NH", "VT"),
  AVG_Rating = c("7", "9", "8", "9")
)
print(doc_df)
```

### Question 2

```{r, q2}
doc_df[1, 2]
doc_df[2:4, ]
doc_df[, 4]
```

### Question 3

```{r, q3}
library(ggplot2)
head(mtcars)


ggplot(mtcars,aes(x=wt, y=mpg, color=factor(cyl))) + geom_point() + labs(x="Weight", y="MPG", title = "Relationship between MPG and Weight")

```

### Question 4

```{r}
# Examine variable summary stats
summary(mtcars)

mpg_weight_cor <- mtcars %>%
select(wt, mpg) %>%
drop_na() %>%
summarize(correlation = cor(wt, mpg))
mpg_weight_cor
```

* It is known that the weight of a vehicle plays a key role in how fuel efficient it is, so both wt and mpg were selected to identify a potential correlation
* The Pearson coefficient is a descriptive statistic that reveals the linear correlation between two variables
* A value between 0 and 1 shows how strong the correlation is, with 1 indicating a strong correlation and 0 an absence of correlation
* The sign of the value indicates the type of correlation (positive or negative)
* In this case, the coefficient value returned is approximately -0.87, which indicates a strong, negative correlation
  * As vehicle weight increases, fuel efficiency (mpg) decreases

---

## Part 2: Practicum Tasks

#### Load data from provided URL

```{r, load_data}

# Documentation from data.world recommends package installation directly from Github
devtools::install_github("datadotworld/data.world-r", build_vignettes = TRUE)

# Load the requisite API token obtained from data.world advanced settings (Thomas's account)
# Original code: token <- readLines('~/RStudioProjects/Summer24_DA5020_Group7_Practicum1/API_token')
# Changed
token <- readLines('API_token')
saved_cfg <- data.world::save_config(token)
data.world::set_config(saved_cfg)

# From data.world R and RStudio integration:
library("data.world")
sql_stmt <- data.world::qry_sql("SELECT * FROM chemical_dependence_treatment_program_admissions_beginning_2007_1")
query_results_df <- data.world::query(
  sql_stmt, "https://data.world/data-ny-gov/ngbt-9rwf")

```

#### Initial evaluation of dataset

-   Determine necessary preparation steps and perform them

-   Discuss distribution, outliers, and prepare summary stats

```{r, q4}

# Evaluate data distribution, outliers, and prepare summary stats

# Reassign query results to more descriptive variable
admissions_data <- query_results_df

# Overview data
glimpse(admissions_data)
summary(admissions_data)

# Visualize outliers in the admissions column
ggplot(admissions_data) +
  labs(title = "Admissions Data Distribution") +
  geom_histogram(mapping = aes(x = admissions), bindwidth = 5)

# Zoom in on low occuring values
ggplot(admissions_data) +
  labs(title = "Admissions Data Distribution (< 8 occurences)") +
  geom_histogram(mapping = aes(x = admissions), binwidth = 5) +
  coord_cartesian(ylim = c(0,8))

# Designate all columns except for year, admissions, and county as categorical (using as.factor())
admissions_data_factors <- admissions_data %>% 
  mutate(
    program_category = as.factor(program_category),
    service_type = as.factor(service_type),
    age_group = as.factor(age_group),
    primary_substance_group = as.factor(primary_substance_group)
  )

# Compute summaries per factor
program_category_summary <- admissions_data_factors %>% 
  group_by(program_category) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  ) 

service_type_summary <- admissions_data_factors %>% 
  group_by(service_type) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
age_group_summary <- admissions_data_factors %>% 
  group_by(age_group) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
primary_substance_group_summary <- admissions_data_factors %>% 
  group_by(primary_substance_group) %>%
  summarize(
    min_admissions = min(admissions),
    median_admissions = median(admissions),
    mean_admissions = mean(admissions),
    max_admissions = max(admissions)
  )
```

```{r}
# Compute outliers for numerical variables
admissions_outliers <- admissions_data_factors %>%
  mutate(
    mean_admissions = mean(admissions, na.rm = TRUE),
    sd_admissions = sd(admissions, na.rm = TRUE)
  ) %>%
  filter(admissions < mean_admissions - 3 * sd_admissions | admissions > mean_admissions + 3 * sd_admissions) %>%
  select(admissions)
admissions_outliers
```

#### Restructure data into appropriate tibbles

```{r county}
# Read in .csv created from https://www.dot.ny.gov/main/business-center/engineering/specifications/local-specifications/county-code as reference
county_codes <- read_csv("county_codes.csv") 
county_codes
```

```{r county1}
county <- admissions_data_factors %>%
  select(county_of_program_location) %>%
  distinct() %>%
  # Join codes with respective counties from county_codes
  left_join(county_codes, by = "county_of_program_location") %>%
  mutate(county_code = case_when(
      # Tagging counties with first two characters and "-NYC"
      county_of_program_location %in% c("Bronx", "Queens", "Kings") ~ paste(str_to_upper(str_sub(county_of_program_location, 1, 2)), "-NYC", sep = ""),
      county_of_program_location == "New York" ~ "NYC",
      # Handle to not treat as NA value, changed code to NS instead
      county_of_program_location == "Nassau" ~ "NS",
      # Handle to not treat as NA value, manually assigned SL 
      county_of_program_location == "St Lawrence" ~ "SL",
      TRUE ~ county_code 
    )
  )
county
```

```{r, program_category}

# Define abbreviations for recoding
program_category_index <- c(
  "Crisis" = "C",
  "Inpatient" = "I",
  "Opioid Treatment Program" = "OTP",
  "Outpatient" = "O",
  "Residential" = "R"
)

# Ceate new column called program_code
admissions_data_coded <- admissions_data_factors %>%
  mutate(program_code = recode(program_category, 
       "Crisis" = "C",
       "Inpatient" = "I",
       "Opioid Treatment Program" = "OTP",
       "Outpatient" = "O",
       "Residential" = "R"))

# Create program_category tibble and recode based on the index directly
program_category_df <- admissions_data_coded %>%
  distinct(program_category, .keep_all = TRUE) %>%
  select(program_code, program_category)

program_category_df

```

```{r, primary_substance_group}
# Sairah
# Define the index for recoding
primary_substance_group_index <- c(
  "Alcohol" = "A",
  "All Others" = "AO",
  "Cocaine incl Crack" = "CC",
  "Heroin" = "H",
  "Marijuana incl Hashish" = "MH",
  "Other Opioids" = "OO"
)

# Update admissions_data_coded with new column called substance_code
admissions_data_coded <- admissions_data_coded %>%
  mutate(substance_code = recode(primary_substance_group, 
        "Alcohol" = "A", 
        "All Others" = "AO",
        "Cocaine incl Crack" = "CC",
        "Heroin" = "H",
        "Marijuana incl Hashish" = "MH",
        "Other Opioids" = "OO"))

# Create Primary Substance Group tibble and recode based on the index directly
primary_substance_group_df <- admissions_data_coded %>%
  distinct(primary_substance_group, .keep_all = TRUE) %>%
  select(substance_code, primary_substance_group)
 
primary_substance_group_df

```

````{r, admissions_data}
# Thomas

# Join county_code data onto main tibble using a full_join by county name
admissions_data_coded_joined <- admissions_data_coded %>% 
  full_join(county, by = "county_of_program_location")

# Final tibble: admissions_data_df
admissions_data_df <- admissions_data_coded_joined %>% 
  select(
    year,
    county_code,
    program_code,
    service_type,
    age_group,
    substance_code,
    admissions
  )

```

#### Define annualAdmissions()

-   Function should derive the total \# of reported admissions per year for the entire state of NY and display these results on a line graph

-   Annotate to show year with highest admissions

-   Explain results

```{r, define_function}
# Thomas -- NEEDS WORK

# This function uses aggregate() to sum the total admissions for every year in the admissions_data_df
# The max point is computed from the aggregated tibble and held for later reference on the graph
# A line graph is prepared using ggplot2 with appropriate labeling

annualAdmissions <- function() {
  # Get aggregated data as its own tibble for easy ref
  total_admissions <- aggregate(admissions_data_df$admissions, 
            by = list(year = admissions_data_df$year), 
            sum) %>% 
    rename(total = x)
  
  # Get maximum
  max_point <- total_admissions[which.max(total_admissions$total), ]
  
  # Plot a line graph
  total_admissions %>% 
    ggplot(mapping = aes(x = year, y = total)) + 
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = 2007:2017) +
    labs(title = "Total admissions per year, all of New York State",
         x = "Total admissions",
         y = "Year") + 
    theme(
      plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
      axis.title.x = element_text(size = 14),
      axis.title.y = element_text(size = 14)
    ) + 
    annotate("text",
             x = max_point$year,
             y = max_point$total,
             label = paste("Max. admissions:", max_point$total),
             vjust = -1)
}

annualAdmissions()

```

#### Analyze % of admissions by county

-   Visualize top 5 counties using a bar chart

-   Explain results

```{r, admissions_by_county}
# Sairah

```

#### Extract various "Rehab" facilities information

-   Use a regex to match all facilities that include the word rehab, rehabilitation, etc.

-   Using filtered data, identify the most prominent (common) substance related to admission for each age group

-   Visualize and explain results

```{r, rehab_regex}
# Caitlin / Thomas / Sairah depending on how it goes
```
